---
title: "Linear models"
---

## Learning objectives

- perform custom tests for model parameters (contrasts, tests of significance);
- interpret and test the significance of interaction terms between continuous and categorical variables, as well as between two categorical variables;


## Content


- Course notes: chapter [4 (Linear regression models)](https://lbelzile.github.io/math60604a/linearmodels.html), Sections 4.5--4.6.


## In class

- [Slides](../slides/MATH60604A-slides5a.html) ([pdf](../slides/MATH60604A-slides5a.pdf) and [code](MATH60604A-slides5a.R))
- [Exercise: interactions](../exercises/interactions.qmd)


## After class 

- Watch the videos of @sec-videos
- [Slides on geometry of OLS](../slides/MATH60604A-slides5b.html) ([pdf](../slides/MATH60604A-slides5b.pdf) and [code](MATH60604A-slides5b.R))
- [Slides on coefficient of determination](../slides/MATH60604A-slides5c.html) ([pdf](../slides/MATH60604A-slides5c.pdf))
- Exercises 2.1, 2.3 and 2.8



## Slides


<p class="text-center"><a class="btn btn-success btn-lg" target="_blank" href="../slides/MATH60604A-slides5a.html">{{< fa arrow-up-right-from-square >}} &ensp;View slides in new window</a> <a class="btn btn-success btn-lg" target="_blank" href="../slides/MATH60604A-slides5a.pdf" role="button">{{< fa file-pdf >}} &ensp;Download PDF of slides</a></p>

<div class="ratio ratio-16x9">
<iframe class="slide-deck" src="../slides/MATH60604A-slides5a.html"></iframe>
</div>


<p class="text-center"><a class="btn btn-success btn-lg" target="_blank" href="../slides/MATH60604A-slides5b.html">{{< fa arrow-up-right-from-square >}} &ensp;View slides in new window</a> <a class="btn btn-success btn-lg" target="_blank" href="../slides/MATH60604A-slides5b.pdf" role="button">{{< fa file-pdf >}} &ensp;Download PDF of slides</a></p>

<div class="ratio ratio-16x9">
<iframe class="slide-deck" src="../slides/MATH60604A-slides5b.html"></iframe>
</div>




<p class="text-center"><a class="btn btn-success btn-lg" target="_blank" href="../slides/MATH60604A-slides5c.html">{{< fa arrow-up-right-from-square >}} &ensp;View slides in new window</a> <a class="btn btn-success btn-lg" target="_blank" href="../slides/MATH60604A-slides5c.pdf" role="button">{{< fa file-pdf >}} &ensp;Download PDF of slides</a></p>

<div class="ratio ratio-16x9">
<iframe class="slide-deck" src="../slides/MATH60604A-slides5c.html"></iframe>
</div>

## Videos {#sec-videos}


```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("files", "youtube-playlist.R"))
playlist_id <- "PLUB8VZzxA8ItMV0PT6lSI6t8Kf-YBPBjj"
slide_details <- tibble::tribble(
~title, ~youtube_id,
"Geometry of least squares", "_amnAMAfvwg",
"Coefficient of determination", "ZFEI7gHJgSY"
)

```


```{r show-youtube-list, echo=FALSE, results="asis"}
youtube_list(slide_details, playlist_id, example = TRUE)
```

