---
title: "Content"
---


The course content is divided in chapters: for each topic,
readers have access to course notes, sample code, online quizzes and exercises.

## Course notes

- [Basics of statistical inference](https://lbelzile.github.io/math60604a/intro.html)
- [Linear regression](https://lbelzile.github.io/math60604a/linear-regression.html)
- [Generalized linear models](https://lbelzile.github.io/math60604a/generalized-linear-models.html)
- [Correlated and longitudinal data](https://lbelzile.github.io/math60604a/correlated-longitudinal-data.html)
- [Linear mixed models](https://lbelzile.github.io/math60604a/linear-mixed-models.html)
- [Survival analysis](https://lbelzile.github.io/math60604a/survival.html)
- [Likelihood](https://lbelzile.github.io/math60604a/likelihood.html)



## Videos and Slides

In this section you can find the links to all videos individually as well as the slides and accompanying code used in the videos.

```{r, echo=FALSE, eval=TRUE, message=FALSE}
source("../files/get_slide_data.R")
knitr::kable(sldat, align = "llccccc")
```

---

## Data

You can access the data sets in **R** by installing the `hecstatmod` package from Github using the following commands (uncomment the first line to install the  `remotes` and `learnr` libraries - do only once):

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("lbelzile/hecstatmod")
remotes::install_github("lbelzile/hecedsm")
```

*Note*: not all data sets are currently documented in `hecstatmod`. Please re-install the package periodically to benefit from the latest features.

## Quiz

The **R** package also contain quizzes: to run the later and launch a new browser window, use the following commands (each line represents a different quiz).

```{r eval = FALSE}
# install.packages("learnr")
learnr::run_tutorial("intro", package = "hecstatmod")
learnr::run_tutorial("likelihood", package = "hecstatmod")
```

